Online supplement accompanying 'The Locally Gaussian Partial Correlation

### 2020-09-09 lgpc_JBES_supp.Rmd
Fixing new round of corrections from Dag

The literature on nonlinear conditional dependence modeling is mostly concerned with the simple case where the vector of conditionally independent variables consists of two scalar random variables, $\Zone = (Z_1, Z_2)$, and this case is therefore the focus of our treatment of the LGPC in the main article. This means that we estimate a single local Gaussian partial correlation function $\widehat \alpha(\z)$. One of the most attractive properties of the LGPC, however, is the close parallel to the multivariate Gaussian distribution, and this is also the case when describing conditional dependence among more than two variables, which in our case leads to the estimation of a *matrix* of local Gaussian partial correlations. Besides needing a slightly more involved notation, this is a straightforward exercise in our framework as we will see below.

@@

All of these parameters can be described and estimated locally in the exact same way as laid out in the main article, resulting from the locally Gaussian joint density $\psi(\z,\fv)$ at the point $\fv = \z$,

@@

\item[\textbf{d)}] \textbf{Let} $\bm{d_1, d_2}$ \textbf{and} $\bm{d_3}$ \textbf{be arbitrary}. This is the general case, and it seems impossible to obtain explicit formulas for the local Gaussian partial correlations, but the general fact still holds of course: The local partial correlation matrix of $\Z_1$and $\Z_2$ given $\Z_3$ can be identified by the lower left hand block matrix of $\D^{-1/2}\fSigma_{12|3}\D^{-1/2}$, and can be estimated using pairwise local correlations in \eqref{eq:matrixdef}.
\end{itemize}

@@

\item[\textbf{(ii)}] If $\bm{d_1 = d_2 = 1}$ \textbf{and} $\bm{d_3 > 1}$, we estimate all correlations in a pairwise manner as laid out in the main article. The limiting distribution of the (scalar) local Gaussian partial correlation is in general given by 
\begin{equation}
\sqrt{nb_n^2}\left(\widehat{\alpha}(z))-\alpha(z)\right)  \stackrel {{\cal L}}{\to} {\cal N}(0,\nabla g(\frho)^{T} {\fLambda} \nabla g (\frho)),
\end{equation}
see eq. (22) in the main article, and see Theorem 2 in Otneim and Tjøstheim (2018) with $\fSigma = \fLambda$ for a definition of $\fLambda$. The computation of $\nabla g$ is outlined in Section \ref{app-asvar}.
\end{itemize}

@@

\begin{itemize}
\item[\textbf{(v)}] In this scheme, we do not use a pairwise set-up for the local correlation only (the one-parameter case), but for all parameters in a bivariate normal distribution; the correlation $\rho(x_1, x_2)$, the expectations $\mu_1(x_1,x_2), \mu_2(x_1,x_2)$ and the variances $\sigma_1(x_1,x_2),\sigma_2(x_1,x_2)$ (the five-parameter case). This setup can also be simplified and fitted to a $p$-variate distribution by estimating $\mu_i = \mu_i(x_1)$ and $\sigma_i = \sigma_i(x_i)$, $i=1,\ldots,p$, marginally. Tjøstheim and Hufthammer (2013) give the asymptotic theory in this case. 
\end{itemize}

Below is a brief summary of the relevant local likelihood theory for the general pairwise case of (ii)--(iv). We also state a general consistency result in Theorem \@ref(thm:A1) at the end of this section.

The pairwise log-likelihood in cases (ii) -- (iv) is given by

\begin{align}
L_{n,b}(\Z_{in}, \Z_{jn}; \rho_b(z_i, z_j)) &= \frac{1}{n}\sum_{t=1}^n K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\psi(v,w;\rho_b(z_i, z_j)) \, \di v \di w,
(\#eq:loglikelihood)
\end{align}
where $\Z_{in} = (Z_{i,1}, Z_{i,2}, \ldots, Z_{i,n})$ and $\Z_{jn} = (Z_{j,1}, Z_{j,2}, \ldots, Z_{j,n})$ are the observation vectors for components $i$ and $j$ respectively, $n$ is the number of observations, $z_i, z_j, v, w \in \mathbb{R}$, $\rho_b = \rho_{ij,b}(z_i,z_j)$ is the local Gaussian correlation parameter, $K_b(v-z_i, w-z_j) = (b_1b_2)^{-1}K(b_1^{-1}(v-z_i))K(b_2^{-1}(w-z_j))$ is a product kernel with bandwidth $b=b_n=(b_1,b_2) = (b_{1n}, b_{2n})$, and $K$ is an non-negative kernel function with compact support such that $\int K(v,w) = 1$. For a fixed bandwidth $b$, $\rho_{0,b}$ is the true local parameter, that is, the minimizer of the local penalty function

\begin{equation}
\int K_b(v-z_i, w-z_j)\big[\psi(v,w; \rho_b(z_i, z_j)) - \log(\psi(v,w; \rho_b(z_i, z_j)))f(v,w)\big] \di v \di w,
(\#eq:penalty)
\end{equation}
where $f$ is the density function of $(Z_i, Z_j)$. Note in particular that $\rho_{0,b} \rightarrow \rho_0$ as $b\rightarrow0$, which is the true local correlation parameter. The reader may consult @tjostheim2013local for more details and regularity conditions in this regard. The corresponding estimate $\widehat \rho = \rho_{n, b}(z_i,z_j)$ is obtained by maximizing the log-likelihood in \@ref(eq:loglikelihood). Note that, writing $\rho$ for $\rho_b$,
\begin{equation}
L_{n,b}(\rho) = \frac{1}{n}\sum_{t=1}^n L^{(b)}(Z_{i,t}, Z_{j,t}; \rho),
\end{equation}
where
\begin{align}
L^{(b)}(Z_{i,t}, Z_{j,t}; \rho) &= K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\rho_b(v,w;\psi_b(z_i, z_j)) \, \di v \di w.
\label{eq:A16}
\end{align}

@@

For fixed bandwidths we have from standard arguments, such as those provided by @hjort1996locally, that 
\begin{equation}
\sqrt{nb^3}\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(0, \Jb^{-1}\Mb(\Jb^{-1})^T),
(\#eq:fixed-h)
\end{equation}
which corresponds to the usual rate for nonparametric density estimation. As $\hh\rightarrow0$, however, we must take extra care when considering the asymptotic covariance matrix. @tjostheim2013local, Section 4,  write the Taylor expansions of $\Mb$ and $\Jb$, being functionals of three variables $(x_1, x_2, x_3)$ or $(z_1,z_2,z_3)$, as
$$\Mb = I_{\bm{M}} + II_{\bm{M}} + o(b^2) \,\,\, \textrm{ and } \Jb = I_{\bm{J}} + II_{\bm{J}} + o(b^2),$$
and consider each term. For example, we write their first term in the expansion of $\Mb$ in the three-variate case as follows:
$$I_{\bm{M}} \sim \int K^2(w_1, w_2, w_3)Ab_wb_w^TA^Tf(\z + b_1w_1 + b_2w_2 + b_3w_3)\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3,$$
where $K$ is a product kernel, and, in our case, $b_w$ is the vector defined by $b_w^T = \begin{pmatrix} 1 & b_1w_1 & b_2w_2 & b_3w_3 \end{pmatrix}$, and $A$ is the $3\times 4$ matrix $A = \begin{pmatrix} v & v_{z_1} & v_{z_2} & v_{z_3}\end{pmatrix}$ with $\fv^T =  \begin{pmatrix} v_1 & v_2 & v_3 \end{pmatrix}$, $v_i = \partial L_n/\partial \rho_i$, where the index $i=1,2,3$ represents the three local correlations, and $v_{z_i} = \partial \fv/\partial z_i$. In the next step, they compute the matrix

@@

$$H_{\z,\hh} = K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right)K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right)v(\Phi^{-1}(\y)),$$
where  $\y = F(\x)$, $y_i = F(x_i)$, $i=1,2$, and we write $K(x_1,x_2) = K(x_1)K(x_2)$. Writing $k(z) = K'(z)$ and $v_{y_i} = \partial v_i/\partial y_i,$
\begin{align*}
\frac{\partial H_{\bm{z},b}}{\partial y_1} 
&= k\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v(\Phi^{-1}(\y))}{b\phi(\Phi^{-1}(y_1))} \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad +  K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v_1(\Phi^{-1}(\y))}{\phi(\Phi^{-1}(y_1))}, \\
\frac{\partial H_{\bm{z},b}}{\partial y_2} 
&= K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) k\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v(\Phi^{-1}(\y))}{b\phi(\Phi^{-1}(y_2))} \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad+  K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v_2(\Phi^{-1}(\y))}{\phi(\Phi^{-1}(y_1))}.
\end{align*}
A typical term in the Taylor expansion of $Y_n(\z)$ takes the following form:

@@

In the original version of this proof we used an argument based on the law of the iterated logarithm of @rio1995functional, leading to a rate of $n^{-1/2}(2 \log \log n )^{1/2}$, but an anynomous referee provided us with a much simpler argument based on a U-statistic decomposition for a strongly mixing process. In the following we omit the index $i=1,2$:

\begin{align*}
\frac{1}{n}\sum_{i=1}^n \left(F_n(X_{i,t}) - F(X_{i,t})\right) &= 
\frac{1}{n^2}\sum_{i=1}^n\sum_{j\neq i}^n \left(I(X_{j, t} \leq X_{i,t}) - F(X_{i,t})\right) + \frac{1}{n^2}\sum_{i=1}^n \left(1 - F(X_{i,t}\right) \\
&= \frac{1}{n} \sum_{i=1}^n \left(0.5 - F(X_{i,t})\right) + O_p\left(n^{-1}\right) \\
&= O_p\left(n^{-1/2}\right) + O_p\left(n^{-1}\right) \\
&= O_p\left(n^{-1/2}\right).
\end{align*}

@@

We have seen that the building blocks of the asymptotic estimation theory of the LGPC between the vectors $\Z_1$ and $\Z_2$ given $\Z_3$ are the pairwise local Gaussian correlations contained in the local correlation matrix $\fSigma$. Let us simply use the notation $\rho_{ij} = \rho_{ij}(z_i, z_j)$ for one of these local correlation and $\hrho_{ij}$ for its estimate as outlined in the main article. If we can prove that the block bootstrap is valid for such a pairwise local likelihood estimate, then it follows fairly straightforwardly that it is also valid for the appropriate vector of such pairwise estimates used in the asymptotic theory laid out above. Note that the 3-coordinate case (i) in Section \@ref(chap:estimation) requires essentially the same arguments.

We can use the bootstrap theory by @lacal2018estimating more or less directly, which in turn is based on Gon\c{c}alves and White [-@gonccalves2002bootstrap; -@gonccalves2004maximum], but we spell out some details  below. First, we need to specify the analogue of the conditions A1--A5 by @lacal2018estimating, where we adapt their notation to our situation by replacing the parameter $\theta$ with $\rho_{ij}$, and the time series $\{X_t, Y_t\}$ by our multivariate observations $\{\Z_t\}$. Since $\rho = \rho_{ij}$ is estimated using the two corresponding data vectors only, we can, for the time being and without loss of generality, assume that the observation series is bivariate: ${\Z_t} = \{Z_{i,t}, Z_{j,t}\}$. Note further that it suffices to consider the true pseudo-observations $\{\Z_t\}$ instead of the empirical pseudo observations $\{\hZ_t\}$ in the following because the difference between them is asymptotically negligible according to Theorem 3.2 in the main article.

@@

\begin{cor}
Let the assumptions of Theorem \ref{thm:E2} hold and let $\fOmega = \fOmega(\frho)$ be the covariance matrix of the $\widehat{\frho}$ vector. Then,
$$
\sqrt{nb_1b_2}(\alpha_{n,b}^*(i,j)-\alpha_{n,b}(i,j)) \buildrel {d_{\mathbb{P}_{\omega}^{*}}} \over {\longrightarrow} {\cal N}(0,\nabla g(\rho)^T \fOmega(\frho)) \nabla g(\rho)) \,\, \mbox{prob}-\mathbb{P}. 
$$ 
\label{cor:E4}
\end{cor}
*Proof:* The proof of Theorem \ref{thm:E2} is given as the proof of Theorem D2 in the online appendix accompanying @lacal2018estimating. This proof can be adapted to prove Corollary \ref{cor:E4}. The starting point consists in replacing $L_{n,b}^*$ and $L_{n,b}$ by $L^* = \sum_{i,j} a_{ij}L_{ij,n,b}^*$ and $L = \sum_{i,j}a_{ij}L_{ij,n,b}$, respectively, where the sum runs over all possible pairs $(i,j)$ corresponding to $\rho_{ij}^{1,2}$, $\rho_{ij}^{1,3}$, $\rho_{ij}^{2,3}$ that are included in the $\frho$-vector. The derivations in  the proof of Theorem D2 by @lacal2018estimating deal with the more complicated five-parameter case. In the present case, however, the vector $\frho$ consists of correlations $\rho_{ij}$ that have been estimated using the one-parameter pairwise setup, where, cf. @otneim2017locally, the quantities $J_{ij}$ and $M_{ij}$ are scalars. Here $L_{ij,n.b}$ is the log likelihood for the pair $(i,j)$ treated in the proof of Theorem D2, and $a_{ij}$ is a corresponding sequence of real numbers. Because of the linearity of the expressions for $L^*$ and $L$ it is easily checked by going step by step that the proof of Theorem D2 of @lacal2018estimating can be carried through for $L^*$ and $L$. The only difference is that the quantities $J_{n,b}$ and $M_{n,b}$ now have to be replaced by $J_{n.b}(a) = \E(-\sum_{i,j}a_{ij}\nabla^2 L_{ij,n,b}(\rho_{0,b}))$ and $\Var(\sqrt{nb_1b_2}\sum_{i,j} a_{ij}\nabla L_{ij,n,b})$. This results in joint asymptotic bootstrap normality using the Cramer-Wold device. Applying the delta rule now proves the Corollary. $\blacksquare$

Let $\M_{ij}$ and $\J_{ij}$ be the $M$-matrix and $J$-matrix appearing in the proof of Corollary \ref{cor:E4} for a fixed $\rho_{ij}$. Note that in the one-parameter case of $\rho_{ij}$ in Corollary \ref{cor:E4}, to first order the covariance matrix $\fOmega$ is a diagonal matrix having ordered values of $M_{ij}/J_{ij}^2$ along the diagonal. This can be seen using the methods of proofs of Theorem 2 of @otneim2017locally and Theorem 2 @otneim2017conditional. Note also that the 3-coordinate case of alternative (i) can be stated and proved in the same way, but a different scaling factor is required to avoid a singular covariance matrix as seen in the proof of Theorem 3.1 in the main article.

@@

*Proof.* The proof follows along the lines of the proof of Theorem \ref{thm:F2}. Corresponding to equation \@ref(eq:t1),
$$
T(\frho_{b,n}) - T(\frho_{b}) \sim \frac{1}{\sqrt{n}}\left\{\int_S h(\frho_b(\z)) \di G_n(\z) + \int_S \nabla_{\frho} h(\frho_b(\z)))^T \fV_n(\z)\right\} \di F(\z)
$$
where $G_n(z) = \sqrt{n}(F_n(z)-F(z))$ and
\begin{align*}
\fV_n(z) &= \sqrt{n} (\frho_{n,b}(\z) - \frho_b(\z)) \\
& \sim \sqrt{n}{\J}_b^{-1}(\z)\left\{\frac{1}{n} \sum_{t=1}^{n}K_b(\Z_t-\z)\fu(\Z_t,\frho_b(\z)) - \int K_b(\fv-\z)\fu(\fv,\frho_b(\z))\psi(\fv,\rho_b(\z)) \di \fv \right\}.
\end{align*}
Here $K_b$ consists of a triple product of kernels $\prod_{i=1}^{3}\frac{1}{b_i}K(\cdot/b_i)$, $\fu(\z,\frho_b(\z)) = \nabla_{\frho}\log \psi(\z,\frho_b(\z))$ where $\psi$ is the density function of a zero mean trivariate normal distribution having a covariance matrix equal to the correlation matrix and with off diagonal elements equal to the elements in $\frho_b(\z)$. This is further developed as in the proof of Theorem \ref{thm:F2}, and the asymptotic normality of $T_{n,b}-T_b$ follows as before from the mixing assumptions and the proof is completed by applying proposition 6.3.9 of @brockwell1991time. $\blacksquare$

@@

# ASYMPTOTIC POWER AND PITMAN EFFICIENCY {#chap:pitman}

>>>>

The literature on nonlinear conditional dependence modeling is mostly concerned with the simple case where the vector of conditionally dependent variables consists of two scalar random variables, $\Zone = (Z_1, Z_2)$, and this case is therefore the focus of our treatment of the LGPC in the main article. This means that we estimate a single local Gaussian partial correlation function $\widehat \alpha(\z)$. One of the most attractive properties of the LGPC, however, is the close parallel to the multivariate Gaussian distribution, and this is also the case when describing conditional dependence among more than two variables, which in our case leads to the estimation of a *matrix* of local Gaussian partial correlations. Besides needing a slightly more involved notation, this is a straightforward exercise in our framework as we will see below.

@@

All of these parameters can be described and estimated locally in the exact same way as laid out in the main article, resulting from the locally approximating Gaussian joint density $\psi(\z,\fv)$ at the point $\fv = \z$,

@@

\item[\textbf{d)}] \textbf{Let} $\bm{d_1, d_2}$ \textbf{and} $\bm{d_3}$ \textbf{be arbitrary}. This is the general case, and it seems impossible to obtain simple explicit formulas for the local Gaussian partial correlations, but the general fact still holds of course: The local partial correlation matrix of $\Z_1$and $\Z_2$ given $\Z_3$ can be identified by the lower left hand block matrix of $\D^{-1/2}\fSigma_{12|3}\D^{-1/2}$, and can be estimated using pairwise local correlations in \eqref{eq:matrixdef}.
\end{itemize}

@@

\item[\textbf{(ii)}] If $\bm{d_1 = d_2 = 1}$ \textbf{and} $\bm{d_3 > 1}$, we estimate all correlations in a pairwise manner as laid out in the main article. The limiting distribution of the (scalar) local Gaussian partial correlation is in general given by 
\begin{equation}
\sqrt{nb_n^2}\left(\widehat{\alpha}(z))-\alpha(z)\right)  \stackrel {{\cal L}}{\to} {\cal N}(0,\nabla g(\frho)^{T} {\fLambda} \nabla g (\frho)),
\end{equation}
see eq. (22) in the main article, and see Theorem 2 in Otneim and Tjøstheim (2018) with $\fSigma = \fLambda$ in that theorem. The computation of $\nabla g$ is outlined in Section \ref{app-asvar}.
\end{itemize}

@@

\begin{itemize}
\item[\textbf{(v)}] In this scheme, we do not use a pairwise set-up for the local correlation only (the one-parameter case), but for all parameters in a bivariate normal distribution; the correlation $\rho(x_1, x_2)$, the expectations $\mu_1(x_1,x_2), \mu_2(x_1,x_2)$ and the variances $\sigma_1(x_1,x_2),\sigma_2(x_1,x_2)$ (the five-parameter case). This setup can also be simplified and fitted to a $p$-variate distribution by estimating $\mu_i = \mu_i(x_i)$ and $\sigma_i = \sigma_i(x_i)$, $i=1,\ldots,p$, marginally. Tjøstheim and Hufthammer (2013) give the asymptotic theory in this case. 
\end{itemize}

Below is a brief summary of the relevant local likelihood theory for the general pairwise case of (ii)--(iv). We also state a general consistency result in Theorem \@ref(thm:A1) at the end of this section. Utilizing Theorem 3.2 in the main article we replace the pseudo normals $\hfZ_t$ of equation (12) of the main paper by vectors $\Z_t$ having standard normal marginals.

The pairwise log-likelihood in cases (ii) -- (iv) can then be written as

\begin{align}
L_{n,b}(\Z_{in}, \Z_{jn}; \rho_b(z_i, z_j)) &= \frac{1}{n}\sum_{t=1}^n K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\psi(v,w;\rho_b(z_i, z_j)) \, \di v \di w,
(\#eq:loglikelihood)
\end{align}
where $\Z_{in} = (Z_{i,1}, Z_{i,2}, \ldots, Z_{i,n})$ and $\Z_{jn} = (Z_{j,1}, Z_{j,2}, \ldots, Z_{j,n})$ are the $\Z$-vectors for components $i$ and $j$ respectively, $n$ is the number of observations, $z_i, z_j, v, w \in \mathbb{R}$, $\rho_b = \rho_{ij,b}(z_i,z_j)$ is the local Gaussian correlation parameter, $K_b(v-z_i, w-z_j) = (b_1b_2)^{-1}K(b_1^{-1}(v-z_i))K(b_2^{-1}(w-z_j))$ is a product kernel with bandwidth $b=b_n=(b_1,b_2) = (b_{1n}, b_{2n})$, and $K$ is an non-negative kernel function with compact support such that $\int K(v,w) = 1$. For a fixed bandwidth $b$, $\rho_{0,b}$ is the true local parameter, that is, the minimizer of the local penalty function

\begin{equation}
\int K_b(v-z_i, w-z_j)\big[\psi(v,w; \rho_b(z_i, z_j)) - \log(\psi(v,w; \rho_b(z_i, z_j)))f(v,w)\big] \di v \di w,
(\#eq:penalty)
\end{equation}
where $f$ is the density function of $(Z_i, Z_j)$. Note in particular that $\rho_{0,b} \rightarrow \rho_0$ as $b\rightarrow0$, which is the true local correlation parameter. The reader may consult @tjostheim2013local for more details and regularity conditions in this regard. The corresponding estimate $\widehat \rho = \rho_{n, b}(z_i,z_j)$ is obtained by maximizing the log-likelihood in \@ref(eq:loglikelihood). Note that, writing $\rho$ for $\rho_b$,
\begin{equation}
L_{n,b}(\rho) = \frac{1}{n}\sum_{t=1}^n L^{(b)}(Z_{i,t}, Z_{j,t}; \rho),
\end{equation}
where
\begin{align}
L^{(b)}(Z_{i,t}, Z_{j,t}; \rho) &= K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\psi(v,w;\rho_b(z_i, z_j)) \, \di v \di w.
\label{eq:A16}
\end{align}

@@

For fixed bandwidths we have from standard arguments, such as those provided by @hjort1996locally, that 
\begin{equation}
\sqrt{nb^3}\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(0, \Jb^{-1}\Mb(\Jb^{-1})^T),
(\#eq:fixed-h)
\end{equation}
which corresponds to the usual rate for nonparametric density estimation. As $\hh\rightarrow0$, however, we must take extra care when considering the asymptotic covariance matrix. @tjostheim2013local, Section 4,  write the Taylor expansions of $\Mb$ and $\Jb$, now being functionals of three variables $(x_1, x_2, x_3)$ or $(z_1,z_2,z_3)$, as
$$\Mb = I_{\bm{M}} + II_{\bm{M}} + o(b^2) \,\,\, \textrm{ and } \Jb = I_{\bm{J}} + II_{\bm{J}} + o(b^2),$$
and consider each term. For example, we write their first term in the expansion of $\Mb$ in the three-variate case as follows:
$$I_{\bm{M}} \sim \int K^2(w_1, w_2, w_3)Ab_wb_w^TA^Tf(\z + b_1w_1 + b_2w_2 + b_3w_3)\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3,$$
where $K$ is a product kernel, and, in our case, $b_w$ is the vector defined by $b_w^T = \begin{pmatrix} 1 & b_1w_1 & b_2w_2 & b_3w_3 \end{pmatrix}$, and $A$ is the $3\times 4$ matrix $A = \begin{pmatrix} v & v_{z_1} & v_{z_2} & v_{z_3}\end{pmatrix}$ with $\fv^T =  \begin{pmatrix} v_1 & v_2 & v_3 \end{pmatrix}$, $v_i = \partial L_n/\partial \rho_i$, where the index $i=1,2,3$ represents the three local correlations, and $v_{z_i} = \partial \fv/\partial z_i$. In the next step, they compute the matrix

@@

$$H_{\z,\hh} = K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right)K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right)v(\Phi^{-1}(\y)),$$
where  $\y = F(\x)$, $y_i = F(x_i)$, $i=1,2$, and we write $K(x_1,x_2) = K(x_1)K(x_2)$. Writing $k(z) = K'(z)$ and $v_{i} = \partial v_i/\partial y_i,$
\begin{align*}
\frac{\partial H_{\bm{z},b}}{\partial y_1} 
&= k\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v(\Phi^{-1}(\y))}{b\phi(\Phi^{-1}(y_1))} \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad +  K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v_1(\Phi^{-1}(\y))}{\phi(\Phi^{-1}(y_1))}, \\
\frac{\partial H_{\bm{z},b}}{\partial y_2} 
&= K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) k\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v(\Phi^{-1}(\y))}{b\phi(\Phi^{-1}(y_2))} \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad+  K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v_2(\Phi^{-1}(\y))}{\phi(\Phi^{-1}(y_2))}.
\end{align*}
A typical term in the Taylor expansion of $Y_n(\z)$ takes the following form:

@@

In the original version of this proof we used an argument based on the law of the iterated logarithm of @rio1995functional, leading to a rate of $n^{-1/2}(2 \log \log n )^{1/2}$, but an anynomous referee provided us with a much simpler argument based on a U-statistic decomposition for a strongly mixing process:

\begin{align*}
\frac{1}{n}\sum_{t=1}^n \left(F_n(X_{i,t}) - F(X_{i,t})\right) &= 
\frac{1}{n^2}\sum_{t=1}^n\sum_{s \neq t}^n \left(I(X_{i,s} \leq X_{i,t}) - F(X_{i,t})\right) + \frac{1}{n^2}\sum_{t=1}^n \left(1 - F(X_{i,t}\right) \\
&= \frac{1}{n} \sum_{t=1}^n \left(0.5 - F(X_{i,t})\right) + O_p\left(n^{-1}\right) \\
&= O_p\left(n^{-1/2}\right) + O_p\left(n^{-1}\right) \\
&= O_p\left(n^{-1/2}\right).
\end{align*}

@@

We have seen that the building blocks of the asymptotic estimation theory of the LGPC between the vectors $\Z_1$ and $\Z_2$ given $\Z_3$ are the pairwise local Gaussian correlations contained in the local correlation matrix $\fSigma$. Let us simply use the notation $\rho_{ij} = \rho_{ij}(z_i, z_j)$ for one of these local correlations and $\hrho_{ij}$ for its estimate as outlined in the main article. If we can prove that the block bootstrap is valid for such a pairwise local likelihood estimate, then it follows fairly straightforwardly that it is also valid for the appropriate vector of such pairwise estimates used in the asymptotic theory laid out above. Note that the 3-coordinate case (i) in Section \@ref(chap:estimation) requires essentially the same arguments.

We can use the bootstrap theory by Lacal and Tjøstheim [-@lacal2017local;-@lacal2018estimating] more or less directly, which in turn is based on Gon\c{c}alves and White [-@gonccalves2002bootstrap; -@gonccalves2004maximum], but we spell out some details  below. First, we need to specify the analogue of the conditions A1--A5 by @lacal2018estimating, where we adapt their notation to our situation by replacing the parameter $\theta$ with $\rho_{ij}$, and the time series $\{X_t, Y_t\}$ by our multivariate observations $\{\Z_t\}$. Since $\rho = \rho_{ij}$ is estimated using the two corresponding data vectors only, we can, for the time being and without loss of generality, assume that the observation series is bivariate: ${\Z_t} = \{Z_{i,t}, Z_{j,t}\}$. Note further that it suffices to consider the true pseudo-observations $\{\Z_t\}$ instead of the empirical pseudo observations $\{\hZ_t\}$ in the following because the difference between them is asymptotically negligible according to Theorem 3.2 in the main article.

@@

\begin{cor}
Let the assumptions of Theorem \ref{thm:E2} hold and let $\fLambda = \fLambda(\frho)$ be the covariance matrix of the $\widehat{\frho}$ vector. Then,
$$
\sqrt{nb_1b_2}(\alpha_{n,b}^*(i,j)-\alpha_{n,b}(i,j)) \buildrel {d_{\mathbb{P}_{\omega}^{*}}} \over {\longrightarrow} {\cal N}(0,\nabla g(\rho)^T \fLambda(\frho)) \nabla g(\rho)) \,\, \mbox{prob}-\mathbb{P}. 
$$ 
\label{cor:E4}
\end{cor}
*Proof:* The proof of Theorem \ref{thm:E2} is given as the proof of Theorem D2 in the online appendix accompanying @lacal2018estimating. This proof can be adapted to prove Corollary \ref{cor:E4}. The starting point consists in replacing $L_{n,b}^*$ and $L_{n,b}$ by $L^* = \sum_{i,j} a_{ij}L_{ij,n,b}^*$ and $L = \sum_{i,j}a_{ij}L_{ij,n,b}$, respectively, where the sum runs over all possible pairs $(i,j)$ corresponding to $\rho_{ij}^{1,2}$, $\rho_{ij}^{1,3}$, $\rho_{ij}^{2,3}$ that are included in the $\frho$-vector. The derivations in  the proof of Theorem D2 by @lacal2018estimating deal with the more complicated five-parameter case. In the present case, however, the vector $\frho$ consists of correlations $\rho_{ij}$ that have been estimated using the one-parameter pairwise setup, where, cf. @otneim2017locally, the quantities $J_{ij}$ and $M_{ij}$ are scalars. Here $L_{ij,n.b}$ is the log likelihood for the pair $(i,j)$ treated in the proof of Theorem D2, and $a_{ij}$ is a corresponding sequence of real numbers. Because of the linearity of the expressions for $L^*$ and $L$ it is easily checked by going step by step that the proof of Theorem D2 of @lacal2018estimating can be carried through for $L^*$ and $L$. The only difference is that the quantities $J_{n,b}$ and $M_{n,b}$ now have to be replaced by $J_{n.b}(a) = \E(-\sum_{i,j}a_{ij}\nabla^2 L_{ij,n,b}(\rho_{0,b}))$ and $\Var(\sqrt{nb_1b_2}\sum_{i,j} a_{ij}\nabla L_{ij,n,b})$. This results in joint asymptotic bootstrap normality using the Cramer-Wold device. Applying the delta rule now proves the Corollary. $\blacksquare$

Let $\M_{ij}$ and $\J_{ij}$ be the $M$-matrix and $J$-matrix appearing in the proof of Corollary \ref{cor:E4} for a fixed $\rho_{ij}$. Note that in the one-parameter case of $\rho_{ij}$ in Corollary \ref{cor:E4}, to first order the covariance matrix $\fLambda$ is a diagonal matrix having ordered values of $M_{ij}/J_{ij}^2$ along the diagonal. This can be seen using the methods of proofs of Theorem 2 of @otneim2017locally and Theorem 2 @otneim2017conditional. Note also that the 3-coordinate case of alternative (i) can be stated and proved in the same way, but a different scaling factor is required to avoid a singular covariance matrix as seen in the proof of Theorem 3.1 in the main article. The 3-coordinate case is in fact closer to the situation in @lacal2018estimating.

@@

*Proof.* The proof follows along the lines of the proof of Theorem \ref{thm:F2}. Corresponding to equation \@ref(eq:t1),
$$
T(\frho_{b,n}) - T(\frho_{b}) \sim \frac{1}{\sqrt{n}}\left\{\int_S h(\frho_b(\z)) \di G_n(\z) + \int_S \nabla_{\frho} h(\frho_b(\z)))^T \fV_n(\z)\right\} \di F(\z)
$$
where $G_n(z) = \sqrt{n}(F_n(z)-F(z))$ and
\begin{align*}
\fV_n(z) &= \sqrt{n} (\frho_{n,b}(\z) - \frho_b(\z)) \\
& \sim \sqrt{n}{\J}_b^{-1}(\z)\left\{\frac{1}{n} \sum_{t=1}^{n}K_b(\Z_t-\z)\fu(\Z_t,\frho_b(\z)) - \int K_b(\fv-\z)\fu(\fv,\frho_b(\z))\psi(\fv,\rho_b(\z)) \di \fv \right\}.
\end{align*}
Here $K_b$ consists of a triple product of kernels $\prod_{i=1}^{3}\frac{1}{b_i}K(\cdot/b_i)$, $\fu(\z,\frho_b(\z)) = \nabla_{\frho}\log \psi(\z,\frho_b(\z))$ where $\psi$ is the density function of a zero mean trivariate normal distribution having a covariance matrix equal to the correlation matrix and with off diagonal elements equal to the elements in $\frho_b(\z)$. This is further developed as in the proof of Theorem \ref{thm:F2}, and the asymptotic normality of $T_{n,b}-T_b$ follows as before from the mixing assumptions and the proof is completed by applying Proposition 6.3.9 of @brockwell1991time. $\blacksquare$

@@

# LOCAL ASYMPTOTIC POWER AND THE PITMAN CRITERION {#chap:pitman}
